---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

The data will be unzipped and read as a data frame. The only processing that is performed is to convert the date column into an actual date field.

```{r, echo=TRUE}
unzip("activity.zip")
rawData <- read.csv("activity.csv")
rawData$date <- as.Date(rawData$date)
```

## What is mean total number of steps taken per day?

First, we examine the number of steps taken per day. In this analysis, missing data will be ignored.

```{r, echo=TRUE}
splitByDay <- split(rawData, rawData$date)
stepsPerDay <- sapply(splitByDay, function(x){ sum(x$steps, na.rm=TRUE)})
hist(stepsPerDay, xlab = "Sum of Steps Per Day", main="Histogram for Steps Per Day")
```

The median steps taken per day is `r median(stepsPerDay)`, while the mean is `r mean(stepsPerDay)`.
## What is the average daily activity pattern

To explore the average daily activity pattern, we take the mean of every five-minute interval across all days.
```{r, echo=TRUE}
splitByInterval <- split(rawData, rawData$interval)
avgSteps <- sapply(splitByInterval, function(x){mean(x$steps, na.rm=TRUE)})
plot.ts(avgSteps,ylab="Average Steps Taken", xlab="Time", main="Average Activity Throughout the Day")

mostActiveInterval <- names(which.max(avgSteps))
hourOfDay <- as.integer(as.integer(mostActiveInterval)/60)
minuteOfHour <-as.integer(mostActiveInterval)- hourOfDay*60
```

On average, the most steps were taking on the interval named `r mostActiveInterval`, which is `r hourOfDay`:`r minuteOfHour`.

## Imputing missing values

```{r, echo=TRUE}
set.seed(1337)

completeCases <-complete.cases(rawData)
completeCaseCount <- sum(completeCases)
missingDataCount <- nrow(rawData) - completeCaseCount

```

There are `r nrow(rawData)` observations in the data set, of which `r missingDataCount` are missing a "steps" count and `r completeCaseCount` are complete.
Since the presence of missing days may introduce some bias, the missing data is filled in by using either the mean or the median value for that interval. Whether the mean or the median is used is determined by a coin flip.

```{r, echo=TRUE}

medianSteps <- sapply(splitByInterval, function(x){median(x$steps, na.rm=TRUE)})
completeData <- rawData

completeData[!completeCases,]$steps <- sapply(rawData[!completeCases,]$interval, function(x){
  coinflip <- rbinom(1,1,.5)[1]
  if(coinflip == 0) {
    avgSteps[as.character(x)]
  } else {
      medianSteps[as.character(x)]
    }
  })

completeSplitByDay <- split(completeData, completeData$date)
completeStepsPerDay <- sapply(completeSplitByDay, function(x){ sum(x$steps, na.rm=TRUE)})
hist(completeStepsPerDay, xlab = "Sum of Steps Per Day", main="Histogram for Steps Per Day With Completed Data")

```

The median steps taken per day is using "complete" data is `r format(median(completeStepsPerDay), digits=10)`, while the mean is `r format(mean(completeStepsPerDay), digits=10)`. The median value has not changed from the original, incomplete data set, but the mean value has gone up by `r mean(completeStepsPerDay) - mean(stepsPerDay) `. The histogram also has a stronger grouping toward the middle.
## Are there differences in activity patterns between weekdays and weekends?

